From physical principles to ???

### 2015-10-24 2015CompleteDerivation/completeDerivation.tex
Using qed symbol instead of box
More work on the assumption rationale

\newenvironment{rationale}{\emph{Rationale}.}{$\Box$}
\newenvironment{justification}{\emph{Justification}.}{$\Box$}
\renewenvironment{proof}{\emph{Proof}.}{$\Box$}

@@

Much of the focus later will be given to the state under deterministic and reversible evolution. However, the non-deterministic part plays a fundamental role as well. In fact: it is this aspect of the evolution that determines what states are available and their description. Suppose we study the motion of a small particle, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. Suppose we study the same small particle as suspended in a fluid, as it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. The issue is that the set of states must be closed under the non-deterministic evolution as well. That is: if we apply the non-deterministic evolution to any state, we still need to end up in one of the possible states. If the Brownian motion is strong enough, the state gets knocked out of the set of states determined by a simple position/momentum pair.

@@

As we have seen, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the individual trajectories are not (those fall within the unstated part). In fact: we cannot assume trajectory and states are defined at all for the unstated part. Consider a muon and its decay products: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the resulting total mass and energy came from the muon. Yet, before the decay, we cannot ascribe an internal state and trajectory to those parts. The state of a muon is not some combination of the state of an electron and two neutrinos.

@@

First, to be able to identify the system, we must first be able to tell it apart from the environment. Intuitively, we can distinguish between two chairs because we can move the first to another room and sit on it without having touched the second. We can manipulate the state of the first system without affecting the second, and vice-versa. So, to tell a system apart from its environment, there has to to be a certain amount of isolation between the two: if I affect its state, the state of the other systems is not affected and vice-versa. This means that the system future and past states are only determined by its own state and the state undergoes deterministic and reversible evolution.

@@

So, if it seems we need to require deterministic and reversible motion, why we call this an assumption? First of all, because it's an idealization: it can never be completely achieved in practice. A system can be prepared or measured up to a certain level of precision. Perfect isolation of a system is impossible both practically (e.g. black-body radiation, gravity, ...) and conceptually (e.g. if the system is perfectly isolated, we cannot interact with it: how can it studied through experimentation?). It's a simplifying assumption that can only be taken if the environment and the internal dynamics of the system interact in such a way that they affect and are little affected by the aspect we are studying. As we saw before, for example, assuming that the states consists of the position and momentum of the center of mass means assuming that the Brownian motion of the body is negligible.

@@

\begin{justification}
The collection of all possible states from a set. The deterministic and reversible evolution maps one element in the set to one and only one element of the set.
\end{justification}

@@

\begin{defn}\label{label}
Let $\mathbbm{I}$ be a set. Let $f_\mathbbm{I} : \mathfrak{S} \rightarrow \mathbbm{I}$ be a function that maps each state to an element of the set. We call the pair $(\mathbbm{I}, f_\mathbbm{I})$ a \emph{state variable}. We call the pair $(i, f_\mathbbm{I}^{-1}(i))$ a \emph{label}, where $i \in \mathbbm{I}$ and $f_\mathbbm{I}^{-1}(i) = \{\mathfrak{s} \in \mathfrak{S} | f_\mathbbm{I}(\mathfrak{s}) = i\}$ is the reverse image of $i$. We call the pair $(I, f_\mathbbm{I}^{-1}(I))$ a \emph{label range}, where $I \subset \mathbbm{I}$.
\end{defn}

@@

We know introduce the idea that a system is decomposable, that is it is made of smaller parts, and we characterize the relationship between the whole and the parts with the following assumption:

\begin{assump}[Infinite reducibility or Classical]\label{classical}
The system is infinitely reducible: it can be thought of as composed of two or more similar but smaller systems, each in its own configuration state, which in turn can be thought of as composed of two or more, ad infinitum. We call \emph{particle} such an infinitesimal part.
\end{assump}

\begin{rationale}
This makes a clear demarcation of what a classical object is (and conversely what a quantum object is not): it is something that can be fully described up to its infinitesimal constituents. This is clearly a \emph{simplifying} assumption, and it is instructive to understand when and why it cannot be used.

The first problem is methodological. As we saw before, we need access to a deterministic and reversible process to be able to study a system and define a state. For two balls, we can imagine studying pieces of the balls in isolation, and then describe the collision between the two by describing what happens at each piece. The classical assumption holds. For an electron and a photon, we cannot take pieces of the electron or the photon, study them in isolation and then describe how each part moves during Compton scattering. The classical assumption does not hold: we do not have suitable physical processes at our disposal.

The second problem is more conceptual. As we saw before, a system cannot be fully isolated from the environment. The smaller the part we study, the more it will be susceptible to fluctuations from the environment, the less we can justify the deterministic assumption, the less we are in a position to define its state. While the overall position of a ball can be considered fairly unaffected by its surrounding, the position of each small piece is going to be constantly affected by air and light scattering on its surface.
\end{rationale}

TODO:

>>>>

\newenvironment{rationale}{\emph{Rationale}.}{\qed}
\newenvironment{justification}{\emph{Justification}.}{\qed}
\renewenvironment{proof}{\emph{Proof}.}{\qed}

@@

Much of the focus later will be given to the state under deterministic and reversible evolution. However, the non-deterministic part plays a fundamental role as well. In fact: it is this aspect of the evolution that determines what states are available and their description. Suppose we study the motion of a cannonball, its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules scatter off its surface randomly, its trajectory is not greatly affect as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory, and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states have. The issue is that the set of states must be closed under the non-deterministic evolution as well. That is: if we apply the non-deterministic evolution to any state, we still need to end up in one of the possible states. If the Brownian motion is strong enough, the state gets knocked out of the set of states determined by a simple position/momentum pair.

@@

As we have seen, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the individual trajectories are not (those fall within the unstated part). In fact: we cannot assume trajectories and states are always defined at all for the unstated part. Consider a muon and its decay into an electron and two neutrinos: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the resulting total mass and energy came from the muon. Yet, before the decay, we cannot ascribe an internal state and trajectory to those parts. The state of a muon is not some combination of the state of an electron and two neutrinos. That is: the unstated part is not just microstates.

@@

First, to be able to identify the system, we must first be able to tell it apart from the environment. Intuitively, we can distinguish between two chairs because we can move the first to another room and sit on it without having touched the second. We can manipulate the state of the first system without affecting the second, and vice-versa. So, to tell a system apart from its environment, there has to to be a certain amount of isolation between the two: if I affect its state, the state of other systems is not affected and vice-versa. This means that the system future and past states are only determined by its own state and the state undergoes deterministic and reversible evolution.

@@

So, if it seems we need to require deterministic and reversible motion, why we call this an assumption? First of all, because it's an idealization: it can never be completely achieved in practice. A system can be prepared or measured up to a certain level of precision. Perfect isolation of a system is impossible both practically (e.g. black-body radiation, gravity, ...) and conceptually (e.g. if the system is perfectly isolated, we cannot interact with it: how can it studied through experimentation?). It's a simplifying assumption that can only be taken if the environment and the internal dynamics of the system interact in such a way that they little affect and are little affected by the aspect we are studying. As we saw before, for example, assuming that the states consists of the position and momentum of the center of mass means assuming that the Brownian motion of the body is negligible.

@@

\begin{justification}
The collection of all possible states form a set. The deterministic and reversible evolution maps one element in the set to one and only one element of the set.
\end{justification}

@@

\begin{defn}\label{label}
Let $\mathbbm{I}$ be a set. Let $f_\mathbbm{I} : \mathfrak{S} \rightarrow \mathbbm{I}$ be a function that maps each state to an element of the set. We call the pair $(\mathbbm{I}, f_\mathbbm{I})$ a \emph{state variable}. We call the pair $(i, f_\mathbbm{I}^{-1}(i))$ a \emph{label}, where $i \in \mathbbm{I}$ and $f_\mathbbm{I}^{-1}(i) = \{\mathfrak{s} \in \mathfrak{S} | f_\mathbbm{I}(\mathfrak{s}) = i\}$ is the inverse image of $i$. We call the pair $(I, f_\mathbbm{I}^{-1}(I))$ a \emph{label range}, where $I \subset \mathbbm{I}$.
\end{defn}

@@

We now introduce the idea that a system is decomposable, that is it is made of smaller parts. To simplify our discussion, we will always assume that our system is homogeneous: all parts can be characterized in the same way, that they are under the same external and internal forces, that the set of states they can occupy is the same. For a non-homogeneous system, we would divide it in homogeneous components and consider the state of all of them. Ultimately, what homogenous means depends on context (e.g. air can be thought as homogeneous if the mixture of gasses does not change in space or in time due to phase transitions or chemical processes) and one must check that such property is conserved by time evolution.

We further characterize the relationship between the whole and the parts with the following assumption:

\begin{assump}[Infinite reducibility or Classical]\label{classical}
The system is infinitely reducible: it can be thought of as composed of two or more similar but smaller systems, each in its own state, which in turn can be thought of as composed of two or more, ad infinitum. We call \emph{particle} such an infinitesimal part. We call \emph{classical} a system for which such an assumption is valid.
\end{assump}

In other words: if we know the state and the evolution for the whole system, we also know the evolution for all its parts; if we know the state and evolution for all the parts, we know the state and the evolution for the whole system. By smaller we do not mean spatial extent, but of smaller amount (e.g. less mass): even at the same spatial location we can imagine that there are infinitely many components (e.g. the mass, as a number, is continuously divisible(.

\begin{rationale}
This is clearly a \emph{simplifying} assumption, and it is instructive to understand when it breaks down.

The first problem is methodological. As we saw before, we need access to a deterministic and reversible process to be able to study a system and define a state. For two balls, we can imagine to isolate some pieces (small enough to be considered infinitesimal in respect to the ball but big enough to contain enough molecules to be considered homogenous) and then describe the collision between the two by describing what happens at each piece. The classical assumption holds. For an electron and a photon, we cannot take pieces of the electron or the photon, study them in isolation and then describe how each part moves during Compton scattering. The classical assumption does not hold: we do not have suitable physical processes at our disposal.

The second problem is more conceptual. As we saw before, a system cannot be fully isolated from the environment. A certain amount of unstated part needs to remain so that the state of the system can be insulated from the external non-deterministic interaction. In practice, the state of each piece of our material is still given by the average of a collection of a large amount of molecules, their respective motion depending not just on the state of the system but on the external environment.  The classical assumption can therefore hold only if we assume that the description of each pieces is, again, not complete. Assuming it at a fundamental level would mean that the motion of each and every molecule of the system is only influenced by the state of the system, no external factor counts. As such, it would be a contradiction with it being a physical system: one we can study experimentally.

While ultimately flawed, the classical assumption can be considered valid for a great number of macroscopic system, and that is why is very useful. One should be cautioned, though, that nowhere in the assumption the spatial extent of the system is mentioned. We may as well have a macroscopic system where a clear independent state cannot be assigned to each part, and the assumption would not hold.
\end{rationale}
